---
title: "Tables and figures in R"
author: "Tham, Wei Yang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document: default
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.asp = 0.618)
```

Good tables and figures are an essential component of every paper. In general, I prefer figures to tables, especially for presenting regression coefficients. I find that most economists use tables too often for my liking, both in papers and presentations. In this document, we'll focus mostly on how to make plots in R with the `ggplot2` package and how to integrate them into a document. We'll use other data-wrangling tools from the `tidyverse` where necessary. We'll do a little bit on tables at the end.

Other resources, some of which I've drawn upon for this workshop:

- Karthik Ram's [slides](https://speakerdeck.com/karthik/introduction-to-ggplot2)
- [R for Data Science](http://r4ds.had.co.nz/data-visualisation.html)
- [Harvard Institute for Quantitative Social Science tutorials](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html#statistical_transformations)

## Why ggplot2?

ggplot2 is based on the grammar of graphics, a phrase coined by Lee Wilkinson. The "grammer" means there is an underlying logic to building a plot. This makes it easier to understand the code used to build a plot, and a lot less effort to memorize arbitrary commands. 

ggplot2 also comes with nice defaults. This means it takes a lot less work get your graphic to a stage where you're satisfied. It's also just a lot more fun to be able to make nice graphs without too much effort. If you want some tips on taking a ggplot2 figure to the next level, check out [this tutorial](https://hopstat.wordpress.com/2016/02/18/how-i-build-up-a-ggplot2-figure/)

## Basics
- `ggplot()` - main function where you specify the dataset and variables to plot
- `geoms` - geometric objects
    + E.g. `geom_point()`, `geom_bar()`, `geom_line()`
- `aes` - aesthetics
    + shape, transparency (alpha), color, fill, linetype
- scales - continuous, discrete, log
- **Side note** - Don't use `qplot()`. Stands for "quick plot", but tricky to add complexity

## Getting started

We'll start with the `gapminder` dataset, which you can get by installing the `gapminder` package. Make sure you have the `tidyverse` packages too.

```{r, eval = FALSE}
install.packages("tidyverse")
install.packages("gapminder")

```

Take a quick peek at the dataset

```{r libs, warning=FALSE, message=FALSE}
library(tidyverse)
library(gapminder)
gapminder
```


### Creating a plot

Try creating your first plot with this code.

```{r, eval = FALSE}
ggplot(gapminder)
```

Ha! Sorry. Try this instead.

```{r}
ggplot(data = filter(gapminder, year == 1997), 
       mapping = aes(x = lifeExp, y = gdpPercap)) + 
  geom_point() 
```

`ggplot()` creates a coordinate system that you can add layers to. To create a graph, you need to specify your

- `data`
- `mapping` - this defines how variables in your dataset are mapped to visual properties (in this case, the x and y axes)

### Points

- Increase the size

```{r}
ggplot(data = filter(gapminder, year == 1997), 
       mapping = aes(x = lifeExp, y = gdpPercap)) + 
  geom_point(size = 5) + 
  ggtitle("For dramatic effect")
```

- Different colours by continent
```{r}
ggplot(data = filter(gapminder, year == 1997), 
       mapping = aes(x = lifeExp, y = gdpPercap, colour = continent)) + 
  geom_point() 
```

Try this:

```{r}
ggplot(data = filter(gapminder, year == 1997), 
       mapping = aes(x = lifeExp, y = gdpPercap)) + 
  geom_point(color = "red") 
```

and this:

```{r}
ggplot(data = filter(gapminder, year == 1997), 
       mapping = aes(x = lifeExp, y = gdpPercap, color = "blue")) + 
  geom_point() 
```

#### Point aesthetics

Different geoms understand different aesthetics. Type `?geom_point` in your console and scroll down to the `Aesthetics` section. The help document says that `geom_point` understands the shape aesthetic, so we can differentiate continents by shapes *and* colours. Shapes are helpful to accommodate black-and-white versions of your paper.

```{r}
ggplot(data = filter(gapminder, year == 1997), 
       mapping = aes(x = lifeExp, y = gdpPercap, shape = continent, color = continent)) + 
  geom_point() 
```



### Lines

Let's look at average worldwide life expectancy over time. We'll calculate mean life expectancy each year and 
then 95\% confidence intervals.

```{r}

lifexp = gapminder %>% 
  group_by(year) %>% 
  summarize(lifexp_mean = mean(lifeExp), lifexp_se = sd(lifeExp)/sqrt(n()))

lifexp = mutate(lifexp, 
                     lifexp_upperci = lifexp_mean + 1.96*lifexp_se, 
                     lifexp_lowerci = lifexp_mean - 1.96*lifexp_se)

lifexp

```

```{r}
ggplot(lifexp, aes(year, lifexp_mean)) + 
  geom_line() 
```

You'll often want to show the confidence intervals e.g. if you have regression coefficients from an event study. 

```{r}
ggplot(lifexp, aes(year, lifexp_mean)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lifexp_lowerci, ymax = lifexp_upperci), alpha = 1/10, fill = "blue")
```

You can also do errorbars.

```{r}
ggplot(lifexp, aes(year, lifexp_mean)) + 
  geom_line() + 
  geom_errorbar(aes(ymax = lifexp_upperci, ymin = lifexp_lowerci), linetype = 2)
```

You can also plot the confidence intervals as lines.

```{r}
ggplot(lifexp, aes(year, lifexp_mean)) + 
  geom_line() + 
  geom_line(aes(year, lifexp_upperci), linetype = 2, color = "red") + 
  geom_line(aes(year, lifexp_lowerci), linetype = 2, color = "red")

```

### Bars

Let's use the `nycflights13` package. Which New York airport had the most number of flights originating from it?

```{r}

# install.packages(nycflights13)
library(nycflights13)

ggplot(flights) + geom_bar(aes(origin))

```

Notice that the counting has been done for you. If you want to display values that are already in the data, use
`geom_col`. 

```{r}
origins = count(flights, origin)

origin_plot = ggplot(origins, aes(origin, n)) + geom_col()
origin_plot

# Same thing
# ggplot(origins, aes(origin, n)) + 
#   geom_bar(stat = "identity")
```

You might want to flip the bars. 

```{r}
origins %>% 
  mutate(origin = reorder(origin, n)) %>% 
  ggplot(aes(origin, n)) + geom_col() + coord_flip()
```


### "3D" plots

ggplot does not have 3D plots, but has geoms for visualizing 3D surfaces in 2D. I often find 2D surfaces to be more interpretable than 3D surfaces anyway. I like heat maps, which you can draw with `geom_raster`. 

```{r}
delays_heatmap = flights %>% 
  group_by(month, day) %>% 
  summarise(dep_delay = mean(dep_delay, na.rm = TRUE), n = n()) %>%  
  ggplot(aes(day, as.factor(month))) + geom_raster(aes(fill = dep_delay)) + 
  ggtitle("Average departure delay from JFK for each month-day of 2013") 

delays_heatmap

```


### Themes

There are many options for deviating from the default grey theme. For papers, I like to use light backgrounds so that color printing is not an issue. 

```{r}
origin_plot + theme_minimal()
```

If you want to change the theme for **all** of your images, use `theme_set()` at the beginning of your code. Copy and paste the code below at the end of the code chunk named `libs` (where we loaded the packages) and knit this `.rmd` file. 

```{r, eval = FALSE}
theme_set(theme_classic())
```

There are themes that come with [ggplot2](http://ggplot2.tidyverse.org/reference/ggtheme.html) and there are also lots other options out there like the `ggthemes` package. 

## Tables

There are a number of helpful tools for making nice tables, but as you've seen with the gapminder dataset, if you're just writing an informal report you can just print the object you want to show and it'll come out pretty neatly. Just to recap:

```{r}
gapminder
```

Another option is `kable` from the `knitr` package. 

```{r}
knitr::kable(head(gapminder)) # without head() will print out entire dataframe
```

`stargazer` is probably your best option for making journal-quality tables. One of its main selling points is that can take regression output directly and convert that into your standard regression table. However, it has some idiosycrasies that can occasionally be inconvenient. Notice below how I have to specify `results = 'asis'` in the chunk options, and `type = "html"` in the function. The latter means that if I want to knit this document as a pdf instead then I have to change the stargazer options. 
<!--
<pre><code>```{r, results = 'asis'}
toymodel = lm(lifeExp ~ gdpPercap, data = gapminder)

stargazer::stargazer(toymodel, type = "html",
 title = "Table with stargazer")
```</code></pre> 
-->

    ```{r, results = 'asis'}`r ''`
    toymodel = lm(lifeExp ~ gdpPercap, data = gapminder)
    
    stargazer::stargazer(toymodel, type = "html", title = "Table with stargazer")
    ```

```{r, results = 'asis'}
toymodel = lm(lifeExp ~ gdpPercap, data = gapminder)

stargazer::stargazer(toymodel, type = "html", title = "Table with stargazer")
```

`stargazer` has a ton of options that you can tweak. The [cheatsheet from Jake Russ (which you can see below)](http://www.jakeruss.com/cheatsheets/stargazer/) is super helpful.

<iframe src = "http://www.jakeruss.com/cheatsheets/stargazer/", style = "width:900px; height:700px"></iframe>

This is also a good opportunity to show you the `broom` package (by David Robinson) for working with regression output. This is one of my favorite packages in R. The `tidy` function converts your regression output into a data frame, making it extremely easy to work with. In particular, it becomes straightforward to plot your coefficients. 

```{r}
toymodel_df = broom::tidy(toymodel)

ggplot(filter(toymodel_df, term == "gdpPercap"), 
       aes(term, estimate, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) + 
  geom_point() + geom_errorbar() + 
  geom_hline(yintercept = 0, linetype = 2) + 
  ggtitle("gdpPercap estimate is statistically significant")
```

Of course, this isn't a very helpful graph since there's only one coefficient, but it's easy to think of situations where it'd be super helpful e.g. event study with different estimates for each point in time.






